\section{Evaluation}
\label{sec:evaluation}

The introduction claimed five contributions.
This section evaluates each against the benchmark evidence.

\textbf{(i) Unified pipeline.}
All six benchmark scenarios (Table~\ref{tab:scenarios}) ran from a single YAML configuration through one binary.
No external tool or intermediate file format was required.
The claim holds.

\textbf{(ii) First genome-wide UMI/duplex simulator.}
The panel UMI scenario (1\,MB, 200x, simplex) completed successfully (Table~\ref{tab:scenarios}).
hg38 chr22 configs for UMI simplex and Twist-style duplex (both at 5x) are included in the hg38 benchmark suite and complete without error.
No other published tool generates genome-wide UMI-tagged reads with configurable family sizes and duplex strand pairing.
The claim holds.

\textbf{(iii) First cfDNA generator with variant support.}
The cfDNA scenario (200x, 2\% purity, cfDNA fragment model) produced reads with nucleosomal fragment distributions and injected somatic variants.
No other tool combines these two features.
The claim holds.

\textbf{(iv) Stochastic VAF.}
All variant-containing benchmarks used per-read Bernoulli sampling.
Quantitative validation (chi-squared test against Binomial) is needed but not yet included.
The claim is supported by design but not yet independently validated.

\textbf{(v) Streaming Rust with bounded memory.}
Per-pair memory converges to ${\sim}0.86$\,KB at high coverage on the 1\,MB reference (Figure~\ref{fig:mem}).
The streaming architecture bounds memory by concurrent batch state rather than total dataset size.
On chr22 at 30x (Table~\ref{tab:hg38}), per-pair memory is similar at ${\sim}0.85$\,KB.
The claim holds in the sense that memory is bounded by channel depth, not dataset size.
The bound itself grows with chromosome length and coverage; the \texttt{output\_buffer\_regions} setting provides a lever to reduce it on memory-constrained systems.

\textbf{Memory limitations.}
Peak RAM for the most demanding configuration tested (chr22, 30x WGS) is 4.11\,GB.
A full human genome at 30x is projected at ${\sim}16$\,GB, which exceeds the 12\,GB benchmark machine and requires either a higher-memory workstation or reducing \texttt{output\_buffer\_regions}.
This is a real constraint for users on budget hardware.
The \texttt{output\_buffer\_regions} knob provides control, but the tradeoff between memory and throughput is not yet characterised across its full range.

\textbf{Thread scaling limitation.}
Adding threads degrades wall time for I/O-bound workloads (1 thread = 140\,s; 12 threads = 160\,s on the 10\,MB / 30x baseline).
FASTQ compression saturates the single writer thread; coordination overhead exceeds compute savings.
For I/O-bound runs, reducing the thread count to 1--2 is the correct setting.
Compute-bound workloads (UMI simulation) benefit from parallelism but were not benchmarked across thread counts.
