\section{Discussion}
\label{sec:discussion}

\varforge{} addresses three critical gaps in the benchmarking ecosystem.

\textbf{UMI/duplex simulation.}
The clinical adoption of error-corrected sequencing has outpaced benchmarking infrastructure.
\varforge{} enables the first systematic evaluation of UMI grouping, consensus calling, and duplex variant detection tools across the full parameter space of family size, UMI error rate, and tumour fraction.

\textbf{cfDNA simulation.}
Liquid biopsy tools such as DELFI and ichorCNA can now be benchmarked against configurable in-silico datasets spanning tumour fractions from 50\% to $<0.01\%$, a regime where wet-lab reference materials are prohibitively expensive.

\textbf{Stochastic VAF.}
By replacing deterministic spike-in with Bernoulli sampling, \varforge{} eliminates the systematic optimism that inflates sensitivity estimates at low VAF and low coverage.

\textbf{Performance for CI/CD.}
The streaming Rust architecture enables integration into automated testing pipelines.
A 1\,MB, 30x simulation completes in ${\sim}15$\,s (baseline, no features).
Panel simulations (1\,MB, 200x, UMI) complete in ${\sim}3$\,min.
Full WGS requires extrapolation, but the linear scaling and bounded memory make it feasible on a standard workstation.

\textbf{Platform-aware error modelling.}
The v0.2.0 sequencing error model layers five passes (cycle-position errors, k-mer context errors, sequencing indels, strand bias, phasing bursts) on top of quality-score generation.
Platform presets for Illumina NovaSeq, PacBio HiFi, and Nanopore R10 supply calibrated defaults.
The model is opt-in: absent configuration preserves backward compatibility.
Combined with the expanded \texttt{learn-profile} subcommand, which now extracts indel rates and per-cycle error rates from BAMs, users can derive data-driven error profiles for any sequencing platform.

\textbf{When threading helps.}
The current benchmark workload (10\,MB, 30x, simple reads) is I/O-bound.
FASTQ compression saturates the single writer thread; adding workers increases coordination overhead without accelerating the bottleneck.
Threading provides a real benefit when the compute cost per read pair rises above the I/O cost.
UMI simulation (PCR family tracking across cycles) and high-mutation-density runs are the primary candidates.
The Panel+UMI scenario (1\,MB, 200x) has higher wall time per base pair than the I/O-bound baseline, confirming that more CPU work per pair shifts the balance.
A parallel BGZF writer or per-worker output files with later merging would make the I/O itself parallelisable, but this is not implemented.
For workloads where thread count does not help, the correct setting is a low thread count to reduce coordination overhead.
The default of all available threads is conservative; users on memory-constrained machines should reduce threads to limit channel buffer size.

Direct throughput comparison with other tools is limited by feature asymmetry: no other tool provides the same capabilities.
BAMSurgeon injects variants at ${\sim}1$ variant per 4--5\,s (${\sim}40$\,min for 500 variants).
\varforge{} applies 500 variants in $<1$\,s within a 140\,s total run (10\,MB, 30x).
ART is single-threaded and lacks variant support.
NEAT is Python-based with limited WGS-scale throughput.
